Post-Disturbance Tree Community Trajectories in a Neotropical Forest

### 2019-03-25 WholePlotTrajectories.tex
relecture

\Abstract{
Understand biodiversity trajectories of tropical forests following
disturbance and clarify the link between biodiversity and disturbance is
urgent to anticipate the impact of the global change context. This calls
to highlight the taxonomic and functional facets of tree community
trajectories and the underlying processes, and to clarify the link
between biodiversity and disturbance, specifically the scope of the
Intermediate Disturbance Hypothesis (IDH) that is debated tropical
forests. In this study, we analyzed tree community response to a
disturbance gradient in a Neotropical forest through diversity,
composition and redundancy post-disturbance trajectories over 30 years.
We considered taxonomic trajectories and functional trajectories based
on 7 leaf, stem and life-history traits. While taxonomic trajectories
differed among communities, functional trajectories remained similar and
convergent, and this decoupling was explained by the variations in
community functional redundancy mitigating the functional impact of
disturbance. We validated the IDH, based on an intensity threshold of
20-25\% of AGB removed, and we clarified its scope, showing that it
applied to taxonomic but not functional diversity, and applied around 20
years after disturbance when the recruitment of pioneers and
late-successionals were balanced. Although consistent, the recovery of
community composition, diversity and redundancy remained unachieved
after 30 years. This acknowledged the need of decades-long cycles
without disturbance to ensure a complete recovery, and questioned
tropical forest community resilience after repeated disturbances.
}

@@

The large areas covered with tropical forests worldwide hold crucial
environmental, economic and social values. They provide wood and
multiple non-timber forest products, shelter a diversified fauna, and
ensure cultural and human well-being. They regulate as well the local
and regional climates, and the carbon, water and nutrient cycles.
However, the growing demand in forests products together with current
global changes increase the pressure on remaining natural forests
\citep{Morales-Hidalgo2015}. These threats affect the natural
disturbance regime that shape and maintain the structure, composition
and functioning of tree communities
\citep{Schnitzer2001, Anderson-Teixeira2013, Sist2015}. It is urgent to
understand community response to disturbance and the underlying
processes to anticipate the fate of tropical forests in the current
context. Disturbance change both community abiotic and biotic
environments, through modifications in the fluxes of light, heat and
water \citep{Goulamoussene2017}. In tropical forests, the forest
environment is maintained after disturbance and there is a suit of
changes in the structure and functioning of communities that consitutes
a secondary succession. For now, this succession has been largely
studied through post-disturbance trajectories of forest structural
parameters such as aboveground biomass, tree height or stem density
\citep{Piponiot2016, Rutishauser2016}. Models based on observed
trajectories then assessed community time to recovery and identified
some of the determinants of post-disturbance trajectories, like forest
structure, forest composition, environmental parameters and disturbance
characteristics \citep{Herault2018}. Regarding community diversity and
composition, however, post-disturbance trajectories have not been as
thoroughly understood \citep{Guitet2018, Molino2001}. Given the variety
of species response to disturbance and the huge diversity of tropical
forests, post-disturbance biodiversity trajectories might be very
diverse \citep{Lindenmayer2012, Garcia_florez2017}. It is essential to
anticipate these trajectories and highlight the link between
biodiversity and disturbance, and specifically how disturbance intensity
and frequency determine the rate, endpoint and duration of
post-disturbance trajectories \citep{Chazdon2003a, Willig2018}.

An early conceptual basis of the linkage between biodiversity and
disturbance is the Intermediate Disturbance Hypothesis (IDH). The IDH
assumes a relationship between community diversity and the intensity and
frequency of disturbance events \citep{Connell1978}. The theory and
states that community diversity peaks at intermediate level of
disturbance. The fluctuations of environmental conditions after
disturbance are assumed to prevent competitive exclusion in fostering
both competitively superior species and fast colonizers
\citep{Shea2004, Pulsford2016}. In tropical forests, however,
observations of the IDH are often blurry compared to theoretical
expectations \citep{Hugues2007, Sheil2003, Norden2017}, and the
validation of the theory remains controversial
\citep{Hubbell2001, Fox2013, Sheil2013}. The link between biodiversity
and disturbance might be complicated by the huge diversity of tropical
communities and a variety of ecological processes may underlie the IDH
(colonisation, facilitation,
coexistence,etc)\citep{Lindenmayer2012, Garcia_florez2017}. There is a
need to clarify the scope of the IDH in terms of the different facets of
community diversity, that are the taxonomic richness, the species
abundance distribution and the community functioning. Besides, there is
a need to delineate the scope of the IDH in time, and highlight the
moment when the conditions for the IDH emerge
\citep{Sheil2003, Shea2004}.

Analysing community response to disturbance requires a set of metrics,
to grasp all aspects of community changes
\citep{Sheil2003, Shea2004, Mayfield2010}. The analysis should first
consider community composition, which is crucial for conservation issues
and reveals the species fostered by the disturbance
\citep{Lavorel2002, Bellwood2006}. Then, diversity metrics encompassing
taxonomic richness and evenness should be considered to assess the
changes in community abundance distribution. A functional approach, that
account for species biological attributes and ecological strategy, would
directly link community diversity, composition and redundancy to
ecosystem functioning and environmental constraints
\citep{Violle2007b, Baraloto2012a}. In that respect, a vast literature
allowed recognizing major traits that represent species ecological
strategy and determine how they respond to changing conditions
\citep{Diaz2005}. Specifically in tropical forests, the functional
approach revealed the post-disturbance deterministic processes entailing
a shift from a dominance of ``conservative'' slow-growing species
dealing with scarce resources, to a dominance of ``acquisitive''
fast-growing species with rapid and efficient use of abundant resources
\citep{Rees2001, Reich2014, Herault2011}. This shift is translated into
the trajectories of average community value of key functional traits
related to resource acquisition (leaf and stem traits) and life-history
strategy (seed mass, maximum size)
\citep{Wright2004, TerSteege2006, Westoby2006a, Chave2009b}. The
functional approach also encompasses the analysis of functional
redundancy, that quantifies the amount of shared trait values among
species \citep{Carmona2016}. The typical high functional redundancy of
hyper-diverse tropical forests \citep{Bellwood2006} mitigates the
impacts of species removal on ecosystem functioning and determines
communities resilience after disturbance \citep{Elmqvist2003, Diaz2005}.

@@

Paracou station in French Guiana (5\textdegree 18'N and
52\textdegree 53'W) is located in a lowland tropical rain forest in a
tropical wet climate with mean annual temperature of 26\textdegree C,
mean annual precipitation averaging 2980 mm.y\textsuperscript{-1} (30-y
period) and a 3-month dry season (\textless{} 100
mm.month\textsuperscript{-1}) from mid-August to mid-November, and a
one-month dry season in March \citep{Wagner2011}. Elevation ranges
between 5 and 50 m and soils correspond to thin acrisols over a layer of
transformed saprolite with low permeability which produces lateral
drainage during heavy rains.

The experiment is a network of 12 6.25 ha plots that underwent a
disturbance gradient of three logging, thinning and fuelwood cutting
treatments (Table \ref{tab:Tab1}) according to a randomized plot design
with three replicate blocks of four plots. The disturbance corresponds
to an intensity gradient. For treatment 1 (T1) 10 trees of commercial
species (of a diameter at 1.3 m height (DBH) equal or above 50 cm) were
felled per hectare. For treatment 2 (T2) 10 trees/ha of commercial
species (DBH \(\geq\) 50 cm) were felled and 30 trees/ha of non-valuable
species (DBH \(\geq\) 40 cm) were removed by poison girdling. For
treatment 3 (T3) 10 trees/ha of commercial species (DBH \(\geq\) 50 cm)
were felled and 30 trees/ha of non-valuable species (15 with DBH
\(\geq\) 50 cm and 15 with DBH \(\geq\) 40 cm) were removed by poison
girdling \citep{Schmitt1990}. Disturbance intensity was measured as the
percentage of aboveground biomass (\%AGB) lost between the first
inventory in 1984 and five years after disturbance \citep{Piponiot2016}
estimated with the BIOMASS R package \citep{Biomass2018}.

@@

Because of the variability in the precision of botanical identification
efforts, we were constraint to conduct the taxonomic composition and
diversity analysis at the genus level. Taxonomic and functional
trajectories of community composition were followed in a two-dimensional
NMDS ordination plan. Two NMDS using abundance-based (Bray-Curtis)
dissimilarity measures were conducted to map either taxonomic or
functional composition, the latter based on the seven leaf, stem and
life history traits (without seed mass classes). Trajectories along time
were reported through the Euclidean distance between the target
inventories and the reference inventories in 1989, \emph{i.e.} 2 years
after disturbance. Univariate trajectories of the leaf, stem and
life-history traits were also visualized with the community weighted
means (CWM) \citep{Diaz2007}. Species seed mass were given in 5 mass
classes. Seed mass trajectories were therefore reported as the
proportion of each class in the inventories (Supplementary materials).

The taxonomic diversity was reported through species richness and the
Hill number translation of the Simpson index \citep{Hill1973}. These
metrics allowed assessing the taxonomic richness as well as evenness,
through the comparison between richness and Simpson diversity. R.esults
will thus be discussed directly in terms of taxonomic richness and
evenness. Both indices belong to the set of HCDT or generalized entropy,
respectively corresponding to the 0 and 2 order of diversity (q),
recommended for diversity studies \citep{Marcon2015b}. The functional
diversity was reported using the functional richness and functional
evenness, \emph{i.e} the Rao index of quadratic entropy. The Rao index
combines species abundance distribution, and the average pairwise
functional dissimilarity between species computed by the gower distance.

@@

Functional redundancy was measured as the overlap among species in
community functional space \citep{Carmona2016}. First, the individuals
of the trait database were mapped in the plane of the two first axes
from a PCA analysis. Then, for each species, the traits probability
density (TPD) were computed through two-dimension kernel density
estimators. Second, for each community, the TDB weigthed by species
abundance were summed accross the functional space. Third, the
functional space was divided into a 100 x 100 grid and the number of
species with a positive TDP was counted in each cell. The average count
across cells minus 1 returnedthe Community Functional Redundancy, which
was the average number of species in the community that share the same
trait values.

@@

\caption{Plot trajectories in terms of taxonomic composition (\textbf{(a)} and \textbf{(c)}) and functional composition (\textbf{(b)} and \textbf{(d)}) in a two-dimensional NMDS plan. Lower panels (\textbf{(c)} and \textbf{(d)}) represent the Euclidean distance to initial condition along the 30 sampled years. Shaded areas are the credibility intervals.}\label{fig:NMDSplans}
\end{figure*}

@@

Community CWM average value of Maximum height at adult stage
(\emph{Hmax}), leaf toughness and wood specific gravity (\emph{WSG})
first decreased and then slightly increased but remained significantly
lower than their initial value (Fig. \ref{fig:CWM}). On the other side,
bark thickness and specific leaf area (\emph{SLA}) increased and while
bark thickness remained substantially high after 30 years, \emph{SLA}
had almost recovered to its initial value. For all traits, the maximum
difference to initial value was correlated to the disturbance intensity.
Positive correlations were observed for Leaf thickness, chlorophyll
content, SLA and bark thickness
(\(\rho_{Spearman}^{Leaf thickness}=0.76\),
\(\rho_{Spearman}^{Chlorophyll content}=0.60\),
\(\rho_{Spearman}^{SLA}=0.93\),
\(\rho_{Spearman}^{Bark thickness}=0.71\)). Negative correlation was
observed for Leaf toughness, WSG and Hmax
(\(\rho_{Spearman}^{Leaf toughness}=-0.53\),
\(\rho_{Spearman}^{WSG}=-0.75\), \(\rho_{Spearman}^{Hmax}=-0.40\)) The
proportions of the three lightest seed mass classes increased in all
disturbed plots, and decreased after 30 years for the lightest class
while it stabilized for the two other (Supp. Mat. - Fig. S2).

@@

The second-degree polynomial regressions between (i) the percentage AGB
loss and (ii) taxonomic and functional diversity after 10, 20 and 30
years best predicted the hump-shaped curve of the disturbance impact
along the disturbance intensity gradient \ref{fig:IDHplot}. The
relationship between the disturbance impact and its intensity was more
markedly hump-shaped for the taxonomic richness than for the Simpson
diversity. For both functional richness and Rao diversity the
relationship was almost linear. The regression model better predicted
the functional richness and Rao diversity
(\(0.55<R^2_{Functional Richness}<0.72\), and
\(0.60<R^2_{Functional Rao}<0.81\)) than the taxonomic richness and
evenness (\(0.21<R^2_{Taxonomic Richness}<0.4\), and
\(-0.15<R^2_{Taxonomic Simpson}<0.43\) respectively).

@@

Regarding functional composition, initial communities were similar and
trajectories were confounded in the functional space. Following
disturbance, the functional composition of pre-disturbance surviving
trees is belived to be the same as this of the initial community
\citep{Herault2018}. Post-disturbance changes in functional composition
then relied upon the recruitment of species or functional types
previously infrequent or absent. These species were probably competitive
pioneers becoming dominant as the most efficient to benefit from the
light, space and nutrients made available by disturbance. This
recruitment of pioneers drove functional composition trajectories
similar for all plots and disturbance intensity. The composition first
went towards more resource-acquisitive strategies, which was translated
by a displacement on the right along the first axis in the functional
plan \citep{Westoby1998, Wright2004, Reich2014}. Thereafter, the
pioneers recruited primarily were progressively excluded by long-lived,
more resistant and shade-tolerant species. The community functional
composition then returned towards more resource-conservative strategies,
suggesting the recovery of the initial community composition and
translated in the functional plan by a displacement left along the first
axis and upward along the second axis (Fig. \ref{fig:NMDSplans}).

@@

Community taxonomic richness and evenness trajectories were determined
by the disturbance intensity, ranging from a limited and temporary
impact to significant and persistant alterations of community diversity.
The taxonomic trajectories markedly changed above an intensity threshold
for which the taxonomic richness was maximized and the taxonomic
evenness remained resilient. The disturbance intensity determined the
balance in the community between pre-disturbance surviving trees, and
trees recruited afterward. Below a 20\%-25\% AGB loss, the trees
surviving after disturbance remained numerous enough to maintain the
high taxonomic richness of the pre-disturbance community
\citep{Bongers2009}. The recruitment of pionners, infrequent or absent
before disturbance, then increased the taxonomic richness all the more
so that the disturbance was intense \citep{Martin2015, Chaudhary2016}.
As these pioneers became more dominant they balanced the usual
hyper-dominance of tropical forests, hence temporarily increasing the
taxonomic evenness \citep{Baraloto2012a}. Above the intensity threshold,
the taxonomic richness of surviving trees was too low to be offset by
the recruitment of pioneers. In the Guiana shield indeed, the pool of
true pioneers specifically recruited after disturbance is restricted to
a few common genera (e.g. \emph{Cecropia} spp., \emph{Vismia} spp.)
\citep{Guitet2018}. The taxonomic richness then decreased following
intense disturbance, all the more so that the disturbance intensity was
high \citep{Molino2001}. At the same time, pioneers became persistently
dominant and prevented a return towards the initial evenness and the
recovery of hyper-dominant shade-tolerant species.

Taxonomic trajectories following disturbance of intermediate intensity
besides highlighted an intermediate time after disturbance, around 15-20
years, for which the taxonomic evenness was maximized. As already
observed in the Guiana shield \citep{Baraloto2012a} and in Borenan
tropical forests \citep{Cannon1998}, taxonomic evenness followed
hump-shaped trajectories with a peak when the recruitment was balanced
between pioneers and late-successional species. This intermediate time
corresponded to the match between the generation time of pionneers,
first to dominate the post-disturbance recruitment, and the time for
competitive exclusion to emerge, which leads to the dominance of
late-successional species.

@@

The decoupling between taxonomic and functional trajectories was
explained by a decrease in the functional redundancy within the
pre-disturbance functional space, due to the loss of species following
disturbance. The recruitment of pioneers, functionally different from
the pre-disturbance functional composition, did not compensate the
decrease of functional redundancy in the first place. Progressively
though, the functional redundancy was restored through the replacement
of the first established species by more competitive, long-lived
pioneers or late-successional species that were functionally closer to
the pre-disturbance community. This replacement followed the lottery
recruitment rules, implying an recruitment easy for the first species
but becoming increasingly hampered by the emergence of interspecific
competition \citep{Busing2002}. The recovery of the functional
redundancy then relied upon the random processe of species recruitment
and was increasingly slow and difficult to anticipate
\citep{Elmqvist2003, Diaz2005}.

@@

AM, EM \& BH designed the study, developed the analysis framework,
interpreted the results and wrote the manuscript. All authors gave final
approval for publication.

>>>>

\Abstract{
Understanding biodiversity trajectories of tropical forests following
disturbance and clarifying the link between biodiversity and disturbance
is urgent to anticipate the impact of the global change context. This
calls to highlight the taxonomic and functional facets of tree community
trajectories and the underlying processes, and to clarify the link
between biodiversity and disturbance, specifically the scope of the
Intermediate Disturbance Hypothesis (IDH) that is debated in tropical
forests. In this study, we analyzed the tree community response to a
disturbance gradient in a Neotropical forest through diversity,
composition and redundancy post-disturbance trajectories over 30 years.
We considered taxonomic trajectories and functional trajectories based
on 7 leaf, stem and life-history traits. While taxonomic trajectories
differed among communities, functional trajectories remained similar and
convergent, and this decoupling was explained by the variations in
community functional redundancy mitigating the functional impact of
disturbance. We validated the IDH, based on an intensity threshold of
20-25\% of AGB removed, and we clarified its scope, showing that it
applied to taxonomic but not functional diversity, and applied around 20
years after disturbance when the recruitment of pioneers and
late-successionals were balanced. Although consistent, the recovery of
community composition, diversity and redundancy remained unachieved
after 30 years. This acknowledged the need of decades-long cycles
without disturbance to ensure a complete recovery, and questioned
tropical forest community resilience after repeated disturbances.
}

@@

The large areas covered with tropical forests worldwide hold crucial
environmental, economic and social values. They provide wood and
multiple non-timber forest products, shelter a diversified fauna, and
ensure cultural and human well-being. They regulate as well the local
and regional climates, and the carbon, water and nutrient cycles.
However, the growing demand in forests products together with current
global changes increase the pressure on remaining natural forests
\citep{Morales-Hidalgo2015}. These threats affect the disturbance regime
that naturally shapes and maintains the structure, composition and
functioning of tree communities
\citep{Schnitzer2001, Anderson-Teixeira2013, Sist2015}. To anticipate
the fate of tropical forests in the current context it is then urgent to
understand the response of communities to disturbance and the underlying
processes. Disturbance impact forest communities in changing both
abiotic and biotic environments, through modifications in the fluxes of
light, heat and water \citep{Goulamoussene2017}. Although in the
tropical context the forest environment is maintained after disturbance,
there is a succession of changes in the structure and functioning of
communities constituting a post-disturbance secondary succession. For
now, this succession has been largely studied through trajectories of
forest structural parameters such as aboveground biomass, tree height or
stem density \citep{Piponiot2016, Rutishauser2016}. Models based on
observed trajectories assessed community time to recovery and identified
some of the determinants of post-disturbance trajectories, like forest
structure, forest composition, environmental parameters and disturbance
characteristics \citep{Herault2018}. Regarding community diversity and
composition however, post-disturbance trajectories have not been as
thoroughly understood \citep{Guitet2018, Molino2001}. Given the variety
of species response to disturbance and the huge diversity of tropical
forests, manifold post-disturbance biodiversity trajectories might
emerge \citep{Lindenmayer2012, Garcia_florez2017}. It is then essential
to anticipate post-disturbance trajectories and highlight the link
between biodiversity and disturbance, specifically the link between
disturbance intensity and frequency and the rate, endpoint and duration
of the trajectories \citep{Chazdon2003a, Willig2018}.

An early conceptual basis of the linkage between biodiversity and
disturbance is the Intermediate Disturbance Hypothesis (IDH). The IDH
assumes a relationship between community diversity and the intensity and
frequency of disturbance events, stating that community diversity peaks
at intermediate level of disturbance \citep{Connell1978}. The theory
assumes that the environmental fluctuations following disturbance foster
both competitively superior species and fast colonizers and thus prevent
competitive exclusion \citep{Shea2004, Pulsford2016}. In tropical
forests, however, observations of the IDH often diverge from theoretical
expectations \citep{Hugues2007, Sheil2003, Norden2017}, and the
validation of the theory remains controversial
\citep{Hubbell2001, Fox2013, Sheil2013}. The link between biodiversity
and disturbance might be complicated by the huge diversity of tropical
communities and a variety of ecological processes may underlie the IDH
(colonisation, facilitation, coexistence,
etc)\citep{Lindenmayer2012, Garcia_florez2017}. There is a need to
clarify the scope of the IDH in terms of the different facets of
community diversity that are taxonomic richness, species abundance
distribution and community functioning. Besides, there is a need to
delineate the scope of the IDH in time, and highlight the moment when
the conditions for the IDH emerge \citep{Sheil2003, Shea2004}.

Analysing community response to disturbance requires a set of metrics,
to grasp all aspects of community changes
\citep{Sheil2003, Shea2004, Mayfield2010}. The analysis should first
consider community composition, which is crucial for conservation issues
and which reveals the species fostered by the disturbance
\citep{Lavorel2002, Bellwood2006}. Then, diversity metrics encompassing
taxonomic richness and evenness should be considered to assess the
changes in community abundance distribution. Besides, a functional
approach that accounts for species biological attributes and ecological
strategy would directly link community diversity, composition and
redundancy to ecosystem functioning and environmental constraints
\citep{Violle2007b, Baraloto2012a}. In that respect, a vast literature
allowed recognizing major traits that represent species ecological
strategy and determine how they respond to changing conditions
\citep{Diaz2005}. Specifically in tropical forests, the functional
approach revealed the post-disturbance deterministic processes entailing
a shift from a dominance of ``conservative'' slow-growing species
dealing with scarce resources, to a dominance of ``acquisitive''
fast-growing species with rapid and efficient use of abundant resources
\citep{Rees2001, Reich2014, Herault2011}. This shift is translated into
the trajectories of average community value of key functional traits
related to resource acquisition (leaf and stem traits) and life-history
strategy (seed mass, maximum size)
\citep{Wright2004, TerSteege2006, Westoby2006a, Chave2009b}. The
functional approach also encompasses the analysis of functional
redundancy, that quantifies the amount of shared trait values among
species \citep{Carmona2016}. The typical high functional redundancy of
hyper-diverse tropical forests \citep{Bellwood2006} mitigates the
impacts of species removal on ecosystem functioning and determines
communities resilience after disturbance \citep{Elmqvist2003, Diaz2005}.

@@

Paracou station in French Guiana (5\textdegree 18'N and
52\textdegree 53'W) is located in a lowland tropical rain forest in a
tropical wet climate with mean annual temperature of 26\textdegree C,
mean annual precipitation averaging 2980 mm.y\textsuperscript{-1} (30-y
period) and a 3-month dry season (\textless{} 100
mm.month\textsuperscript{-1}) from mid-August to mid-November, and a
one-month dry season in March \citep{Wagner2011}. The mean annual
temperature is 26\textdegree C. Elevation ranges from 5 to 50 m. Across
all plots the topography mainly corresponds to hilltops or hillsides,
while bottomlands cover less than 1 \% of the area. Plots are shallow
ferralitic acrisols over a layer of transformed saprolite with low
permeability and lateral drainage. Soil conditions are homogeneous, to
the exception of the highest hilltops where the thick surface allows a
free vertical drainage \citep{Gourlet-Fleury2004}.

The experiment is a network of twelve 6.25 ha plots (Table
\ref{tab:Tab1}) that underwent three disturbance treatments in 1987
according to a randomized plot design \citep{Gourlet-Fleury2004}.\\
The experiment comprised three replicates of three sylvicultural
treatments (hereafter plots T1, T2 and T3) and three control plots (T0).
All treatments T1, T2 and T3 comprised the logging of 10 trees/ha with
50 cm minimum DBH that belonged to a set of 58 commercial species
\citep{Gourlet-Fleury2004}. Treatment T2 additionally comprised a
thinning treatment by poison-girdling of non-commercial, randomly
selected species with an average 30 trees/ha with 40 cm minimum DBH.
Treatment T3 additionally comprised the logging of 15 trees /ha with 40
cm minimum DBH and the poison-gidling of 20 trees/ha with a 50 cm
minimum DBH, all belonging to non-commercial species. Considering the
silvicultural treatments and the following damage, disturbance intensity
was measured as the percentage of aboveground biomass (\%AGB) lost
between the first inventory in 1984 and five years after disturbance
\citep{Piponiot2016} estimated with the BIOMASS R package
\citep{Biomass2018}. The three treatments constituted a disturbance
intensity gradient with increasing of above-ground biomass (AGB) lost
and surface disturbed.

@@

Because of the variability in the precision of botanical identification
efforts, we were constrained to conduct the taxonomic composition and
diversity analysis at the genus level. Taxonomic and functional
trajectories of community composition were followed in a two-dimensional
NMDS ordination plane. Two NMDS using abundance-based (Bray-Curtis)
dissimilarity measures were conducted to map either taxonomic or
functional composition, the latter based on the seven leaf, stem and
life history traits (without seed mass classes). Trajectories along time
were reported through the Euclidean distance between the target
inventories and the reference inventories in 1989, \emph{i.e.} 2 years
after disturbance. Univariate trajectories of the leaf, stem and
life-history traits were also visualized with the community weighted
means (CWM) \citep{Diaz2007}. Species seed mass were given in 5 mass
classes, and seed mass trajectories were reported as the proportion of
each class in the inventories (Supplementary materials).

The taxonomic diversity was reported through species richness and the
Hill number translation of the Simpson index \citep{Hill1973}. These
metrics allowed assessing the taxonomic richness as well as evenness,
through the comparison between these two metrics.: results will be
discussed directly in terms of taxonomic richness and evenness. Both
indices belong to the set of HCDT or generalized entropy, respectively
corresponding to the 0 and 2 order of diversity (q), recommended for
diversity studies \citep{Marcon2015b}. The functional diversity was
reported using the functional richness and functional evenness,
\emph{i.e} the Rao index of quadratic entropy. The Rao index combines
species abundance distribution, and the average pairwise functional
dissimilarity between species computed by the gower distance.

@@

Functional redundancy was measured as the overlap among species in
community functional space \citep{Carmona2016}. First, the individuals
of the trait database were mapped in the plane of the first two axes
from a PCA analysis, which lowered the weight of correlations among
traits as the PCA axes were combinations of most decoupled traits. For
each species, the traits probability density (TPD) were computed from
the mapping of individuals through two-dimension kernel density
estimators. Second, for each community, the TPD weigthed by species
abundance were summed accross the functional space. Third, the
functional space was divided into a 100 x 100 grid and the number of
species with a positive TPD was counted in each cell. The average count
across cells minus 1 returned the Community Functional Redundancy, which
was the average number of species in the community that share the same
trait values.

@@

\caption{Plot trajectories in terms of taxonomic composition (\textbf{(a)} and \textbf{(c)}) and functional composition (\textbf{(b)} and \textbf{(d)}) in a two-dimensional NMDS plane. Lower panels (\textbf{(c)} and \textbf{(d)}) represent the Euclidean distance to initial condition along the 30 sampled years. Shaded areas are the credibility intervals.}\label{fig:NMDSplans}
\end{figure*}

@@

Community CWM average value of Maximum height at adult stage
(\emph{Hmax}), leaf toughness and wood specific gravity (\emph{WSG})
first decreased and then slightly increased but remained significantly
lower than their initial value (Fig. \ref{fig:CWM}). On the other side,
bark thickness and specific leaf area (\emph{SLA}) increased and while
bark thickness remained substantially high after 30 years, \emph{SLA}
had almost recovered to its initial value. For all traits, the maximum
difference to initial value was correlated to the disturbance intensity.
Positive correlations were observed for Leaf thickness, chlorophyll
content, SLA and bark thickness
(\(\rho_{Spearman}^{Leaf thickness}=0.76\),
\(\rho_{Spearman}^{Chlorophyll content}=0.60\),
\(\rho_{Spearman}^{SLA}=0.93\),
\(\rho_{Spearman}^{Bark thickness}=0.71\)). Negative correlation was
observed for Leaf toughness, WSG and Hmax
(\(\rho_{Spearman}^{Leaf toughness}=-0.53\),
\(\rho_{Spearman}^{WSG}=-0.75\), \(\rho_{Spearman}^{Hmax}=-0.40\)) The
proportions of the three lightest seed mass classes increased in all
disturbed plots. After 30 years the proportion of lightest seed mass
class decreased while it stabilized for the two other (Supp. Mat. - Fig.
S2).

@@

The second-degree polynomial regressions between (i) the percentage AGB
loss and (ii) taxonomic and functional diversity after 10, 20 and 30
years best predicted the hump-shaped curve of the disturbance impact
along the disturbance intensity gradient (Fig. \ref{fig:IDHplot}). The
relationship between the disturbance impact and its intensity was more
markedly hump-shaped for the taxonomic richness than for the Simpson
diversity. For both functional richness and Rao diversity the
relationship was almost linear. The regression model better predicted
the functional richness and Rao diversity
(\(0.55<R^2_{Functional Richness}<0.72\), and
\(0.60<R^2_{Functional Rao}<0.81\)) than the taxonomic richness and
evenness (\(0.21<R^2_{Taxonomic Richness}<0.4\), and
\(-0.15<R^2_{Taxonomic Simpson}<0.43\) respectively).

@@

Regarding functional composition, initial communities were similar and
trajectories were confounded in the functional space. Following
disturbance, the functional composition of pre-disturbance surviving
trees is believed to be the same as this of the initial community
\citep{Herault2018}. Post-disturbance changes in functional composition
then relied upon the recruitment of species or functional types
previously infrequent or absent. These species were probably competitive
pioneers becoming dominant as the most efficient to benefit from the
light, space and nutrients made available by disturbance. This
recruitment of pioneers drove functional composition trajectories
similar for all plots and disturbance intensity. The composition first
went towards more resource-acquisitive strategies, which was translated
by a displacement to the right along the first axis in the functional
plane \citep{Westoby1998, Wright2004, Reich2014}. Thereafter, the
pioneers recruited primarily were progressively excluded by long-lived,
more resistant and shade-tolerant species. The community functional
composition then returned towards more resource-conservative strategies,
suggesting the recovery of the initial community composition and
translated in the functional plane by a displacement left along the
first axis and upward along the second axis (Fig. \ref{fig:NMDSplans}).

@@

Community taxonomic richness and evenness trajectories were determined
by the disturbance intensity, ranging from a limited and temporary
impact to significant and persistant alterations of community diversity.
The taxonomic trajectories markedly changed above an intensity threshold
for which the taxonomic richness was maximized and the taxonomic
evenness remained resilient. The disturbance intensity determined the
balance in the community between pre-disturbance surviving trees, and
trees recruited afterward. Below a 20\%-25\% AGB loss, the trees
surviving after disturbance remained numerous enough to maintain the
high taxonomic richness of the pre-disturbance community
\citep{Bongers2009}. The recruitment of pioneers, infrequent or absent
before disturbance, then increased the taxonomic richness all the more
so that the disturbance was intense \citep{Martin2015, Chaudhary2016}.
As these pioneers became more dominant they balanced the usual
hyper-dominance of tropical forests, hence temporarily increasing the
taxonomic evenness \citep{Baraloto2012a}. Above the intensity threshold,
the taxonomic richness of surviving trees was too low to be offset by
the recruitment of pioneers. In the Guiana shield indeed, the pool of
true pioneers specifically recruited after disturbance is restricted to
a few common genera (e.g. \emph{Cecropia} spp., \emph{Vismia} spp.)
\citep{Guitet2018}. The taxonomic richness then decreased following
intense disturbance, all the more so that the disturbance intensity was
high \citep{Molino2001}. At the same time, pioneers became persistently
dominant and prevented a return towards the initial evenness and the
recovery of hyper-dominant shade-tolerant species.

Taxonomic trajectories following disturbance of intermediate intensity
besides highlighted an intermediate time after disturbance, around 15-20
years, for which the taxonomic evenness was maximized. As already
observed in the Guiana shield \citep{Baraloto2012a} and in Bornean
tropical forests \citep{Cannon1998}, taxonomic evenness followed
hump-shaped trajectories with a peak when the recruitment was balanced
between pioneers and late-successional species. This intermediate time
corresponded to the match between the generation time of pionneers,
first to dominate the post-disturbance recruitment, and the time for
competitive exclusion to emerge, which leads to the dominance of
late-successional species.

@@

The decoupling between taxonomic and functional trajectories was
explained by a decrease in the functional redundancy within the
pre-disturbance functional space, due to the loss of species following
disturbance. The recruitment of pioneers, functionally different from
the pre-disturbance functional composition, did not compensate the
decrease of functional redundancy in the first place. Progressively
though, the functional redundancy was restored through the replacement
of the first established species by more competitive, long-lived
pioneers or late-successional species that were functionally closer to
the pre-disturbance community. This replacement followed the lottery
recruitment rules, implying an easy recruitment for the first species
but becoming increasingly hampered by the emergence of interspecific
competition \citep{Busing2002}. The recovery of the functional
redundancy then relied upon the random processe of species recruitment
and was increasingly slow and difficult to anticipate
\citep{Elmqvist2003, Diaz2005}.

@@

AM, EM \& BH designed the study, developed the analysis framework and
interpreted the results. AM wrote the manuscript with contributions by
EM \& BH. All authors gave final approval for publication.
